---
title: "Time Series Forecasting: Simple Smoothing Methods"
author: "Xiaorui (Jeremy) Zhu"
date: "02/15/2026"
output:
  slidy_presentation:
    font_adjustment: 2
    footer: Copyright (c) 2026, Xiaorui (Jeremy) Zhu
  beamer_presentation: default
  ioslides_presentation: default
editor_options:
  chunk_output_type: console
---

# Case: Forecasting Apple Revenue 1 {.bigger .incremental}

Apple Inc. is a global technology company with rapidly changing revenue driven by product launches and market trends.

- In 2020, Apple reported $64.7 billion in quarterly revenue, boosted by strong demand during the COVID-19 pandemic.

- Revenue fluctuates over time due to factors like seasonality, new products, and economic conditions.

# Case: Forecasting Apple Revenue 2 {.bigger .incremental}

Key Question:

1. How can we use past revenue data to understand patterns and forecast future performance?

2. As an analyst, you will apply time-series models to analyze and predict Apple’s revenue.

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(readxl)
# Load data
data <- read_excel("data/Revenue_Apple.xlsx")
# View first few rows
data[1:2,]

library(ggplot2)

data$Time <- 1:nrow(data)

ggplot(data, aes(x = Time, y = Revenue)) +
  geom_line(size = 1) +
  labs(title = "Apple Revenue Over Time",
       x = "Time",
       y = "Revenue") +
  theme_minimal()
```


# What is Time-series Data and Forecasting?

- A time series is data collected over time (e.g., quarterly revenue)
- Time periods can be expressed as year, quarter, month, week, day, hour or seconds.
- Goal: use past data to predict future values
- Important for business decisions (planning, budgeting)

# Time-series Data

A time series is characterized by four components:

1. **Trend**: long-term up/down movements

2. **Seasonal**: repetitions over a one-year period, magnitudes of the up/down swings are more predictable

3. **Cyclic**: wavelike fluctuations or business cycles, duration lasts several years and differs from cycle to cycle

4. **Random**: unexplained movements

# Simple Smoothing Methods

---

## Why Do We Need Smoothing?

- Real-world data fluctuates (noise)
- Hard to see underlying patterns
- Smoothing helps:
  - Reveal trends
  - Improve forecasts

---

## Moving Average Method

**Idea:** Use the average of the last *n* observations

$$
F_{t+1} = \frac{Y_t + Y_{t-1} + ... + Y_{t-n+1}}{n}
$$

- Simple and intuitive
- Larger window (n) → smoother but slower to react

---

## Moving Average Example (R)

```{r}
library(readxl)
# Load Apple revenue data
data <- read_excel("data/Revenue_Apple.xlsx")
table(data$Year)
# Example: suppose data starts at 2005 Q1
ts_revenue <- ts(data$Revenue, 
              start = c(2010, 1),   # Year, Quarter
              frequency = 4)        # Quarterly data

# Plot raw data
plot(ts_revenue, type="l", main="Apple Revenue", ylab="Revenue")
# Moving average (3-period)
ma <- stats::filter(ts_revenue, rep(1/3, 3), sides=1)
lines(ma, col="blue", lwd=2)
legend("topleft", legend=c("Actual","Moving Avg"), col=c("black","blue"), lty=1)
```

---

## Simple Exponential Smoothing (SES)

**Idea:** Weighted average of past values

$$
F_{t+1} = \alpha Y_t + (1 - \alpha)F_t
$$

- $\alpha$ controls how fast the model reacts
- More flexible than moving average

---

## How Simple Exponential Smoothing (SES) Works

- Instead of averaging equally (like moving average), SES:
    - Gives more importance to recent data
    - Gradually discounts older observations
    
- Forecast is updated each period using:

$$
F_{t+1} = \alpha Y_t + (1 - \alpha)F_t
$$

- New forecast = (weight × latest actual) + (remaining weight × previous forecast)

---

## Exponential Smoothing Example (R)

```{r}
library(forecast)
# use the time series date to Fit SES model
ses_model <- ses(ts_revenue)
# Plot
plot(ses_model, main="SES Forecast for Apple Revenue")
```

---

## Exponential Smoothing Forecasting

```{r}
forecast_values <- forecast(ses_model, h=4)
plot(forecast_values)
```

---

## Autoregressive (AR) Model

**Idea:** Use past values to predict future values

$$
Y_t = c + \phi_1 Y_{t-1} + \phi_2 Y_{t-2} + ... + \epsilon_t
$$

- Uses correlation in time series
- More advanced than smoothing

---

## AR Model Example (R)

```{r}
# Step 1: Split the training and testing sets
train <- window(ts_revenue, end = c(2018,4))   # adjust
test  <- window(ts_revenue, start = c(2019,1))

# Step 2: Fit model on training
library(forecast)
model <- Arima(train, order = c(1,0,0))
fc <- forecast(model, h = length(test))

# Step 3: Evaluate accuracy
accuracy(fc, test)
```

---

## Optional Visualization

```{r}
plot(fc)
lines(test, col="red")
legend("topleft", 
       legend=c("Forecast","Actual"), 
       col=c("blue","red"), 
       lty=1)
```

---


## Better Model and Visualization

```{r}
# Better model 
model2 <- auto.arima(train)
fc2 <- forecast(model2, h = length(test))
# Visualization
plot(fc2)
lines(test, col="red")
legend("topleft", 
       legend=c("Forecast","Actual"), 
       col=c("blue","red"), 
       lty=1)
```

---
## Key Takeaways

- Moving Average: simple and stable
- Exponential Smoothing: more flexible
- AR Model: uses time dependence
- Start simple, then move to more advanced models
