---
title: "Data Wrangling Examples in R"
author: "Xiaorui (Jeremy) Zhu"
date: "01/25/2026"
output:
  slidy_presentation:
    font_adjustment: +2
    footer: "Copyright (c) 2025, Xiaorui (Jeremy) Zhu"
    # df_print: paged
editor_options: 
  chunk_output_type: console
---


```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, warning=FALSE, message = FALSE)
```

# Aircraft data 

The aircraft.xlsx dataset contains columns for MD ([United States Air Force (USAF)](https://en.wikipedia.org/wiki/List_of_active_United_States_Air_Force_aircraft) ), FH (Flight Hours), FY (Fiscal Year), Cost, and Gallons. 

## Import 

[aircraft.xlsx](https://xiaorui.site/Data-Mining-Business-Analytics/lab/data/aircraft.xlsx)

1. What spreadsheets are contained in the aircraft.xlsx file?
2. Read in the Trainers worksheet data without the header
information.

# Q1 Solution

1. What spreadsheets are contained in the aircraft.xlsx file?
```{r}
library(readxl)
excel_sheets("data/aircraft.xlsx")
```

# Q2 Solution

2. Read in the Trainers worksheet data without the header information.
```{r}
aircraft <- read_excel("data/aircraft.xlsx", sheet = "Trainers", skip = 3)
head(aircraft)
```


# Numeric Understanding of data: Question 3 to 7

3. Look at the summary information for this data. Anything seem odd?

4. Which aircraft MDs are represented?

5. Are there any missing years between 1996-2014 in this data?

6. 90% of flying hours fall under what value?

7. What is the spread of the range of costs?

# Q3 Solution

3. Look at the summary information for this data. Anything seem odd??

```{r}

```

# Q4 Solution

4. Which aircraft MDs are represented?
```{r}

```

# Q5 Solution

5. Are there any missing years between 1996-2014 in this data?

```{r}

```

# Q6 Solution

6. 90% of flying hours fall under what value?

```{r}

```

# Q7 Solution

7. What is the spread of the range of costs?

```{r}

```

[go to top](#header)

# Visual Understanding of data: Question 8 to 10
8. How would you describe the distribution of flying hours?

9. If we wanted to focus on only the trainers with the largest variance in flying hours, which MDs would we select?

10. Are all FYs equally represented?

# Q8 Solution

8. How would you describe the distribution of flying hours?

```{r}

```

# Q9 Solution

9. If we wanted to focus on only the trainers with the largest variance in flying hours, which MDs would we select?

```{r}

```

# Q10 Solution

10. Are all FYs equally represented?

```{r}

```

[go to top](#header)   

[go back to course website](https://xiaorui.site/Data-Mining-Business-Analytics/)


