---
title: "Data Wrangling Examples in R"
author: "Xiaorui (Jeremy) Zhu"
date: "01/08/2026"
output:
  slidy_presentation:
    font_adjustment: +2
    footer: "Copyright (c) 2025, Xiaorui (Jeremy) Zhu"
    # df_print: paged
editor_options: 
  chunk_output_type: console
---


```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, warning=FALSE, message = FALSE)
```

# Aircraft data 

## Import 

[aircraft.xlsx](https://xiaorui.site/Data-Mining-Business-Analytics/lab/data/aircraft.xlsx)

1. What spreadsheets are contained in the aircraft.xlsx file?
2. Read in the Trainers worksheet data without the header
information.

# Q1 Solution

1. What spreadsheets are contained in the aircraft.xlsx file?
```{r}
library(readxl)
excel_sheets("data/aircraft.xlsx")
```

# Q2 Solution

2. Read in the Trainers worksheet data without the header information.
```{r}
aircraft <- read_excel("data/aircraft.xlsx", sheet = "Trainers", skip = 3)
head(aircraft)
```


# Question 3 to 7

3. Look at the summary information for this data. Anything seem odd?

4. Which aircraft MDs are represented?

5. Are there any missing years between 1996-2014 in this data?

6. 90% of flying hours fall under what value?

7. What is the spread of the range of costs?

# Q3 Solution

3. Look at the summary information for this data. Anything seem odd??

```{r}
summary(aircraft)
```

# Q4 Solution

4. Which aircraft MDs are represented?
```{r}
table(aircraft$MD)
```

# Q5 Solution

5. Are there any missing years between 1996-2014 in this data?

```{r}
unique(aircraft$FY)

table(aircraft$FY)
```

# Q6 Solution

6. 90% of flying hours fall under what value?

```{r}
quantile(aircraft$FH, prob = .9)
```

# Q7 Solution

7. What is the spread of the range of costs?

```{r}
range(aircraft$Cost)
```


[go to top](#header)